---
title: 'U.S. Tax Reform:  A Look at the Flat Tax Model'
author: "Cory R. Robinson"
date: "August 5, 2015"
output: html_document
runtime: shiny
---
```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(XLConnect)
library(ggplot2)
library(reshape2)
library(scales)
library(foreach)
library(doParallel)
library(geiger)
library(sp)
library(rgeos)
library(Bolstad2)
library(knitr)
library("dplyr")
setwd("~/Research/Tax/data")
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, 
                      cache=TRUE, fig.width=20, fig.height=12)

# load the data
df <- readWorksheetFromFile("clean2012.xls", 
                            sheet=1, 
                            startRow = 2,
                            endCol = 40)
df <- data.frame(df)
df <- df[-1,]
```

With the 2016 U.S. Presidential Election under way, one thing that many candidates have already begun talking about is tax reform.  During my brief time as a tax preparer, I barely scratched the surface of the tax policy as I calculated capital gains to be taxed and combed through the list of itemized deductions that could be taken in the thousand or so pages of the Master Tax Guide.  Fortunately, I didn't have to look at the daunting 74,608 page long U.S. Federal Tax Code!  Admittedly, a tax code this long is quite a complex machine filled with many different ways of taxing and crediting.  Taxes are one of the necessities of life that we have all grown to love to hate.  As complex a machine that the tax code is, people have and will continue to find loopholes in the code in an attempt to pay the least amount of taxes possible by deducting everything they can, including the kitchen sink!  

Thus far in the 2016 race, the proposed tax reform plans range from re-writing a few pages of the tax code to completely re-writing the entire tax code and eliminating several thousand pages of tax code garbage.  Here are some ideas of a few of the candidates in the 2016 race:

* **Hillary Clinton** - rewriting a few pages to include an increase in short-term capital gains taxes along with introducing some tax credits related to profit sharing, apprenticeships, and renewable energy.
* **Ben Carson** - rewriting entire tax code and only implementing a 10% flat tax rate.
* **Rand Paul** - rewriting most of the tax code and implementing a 14.5% flat tax rate, with some credits and deductions similar to those in the current tax code.
* **Mike Huckabee** - rewriting entire tax code and proposing the Fair Tax model, which essentially eliminates the income tax, and the only tax that anyone will pay is a sales tax that is larger than any sales taxes that we currently have.

Since my focus here is to inform voters of the mechanics of an extensive tax reform, let's take a closer look at the various flat tax models.

First off, I would like to note that there are [several countries](https://en.wikipedia.org/wiki/Flat_tax#Around_the_world) around the world, and even some of our own [U.S. states](https://en.wikipedia.org/wiki/Flat_tax#At_the_state_or_provincial_level) that have implemented successful flat tax models[^1].

**True Flat Tax**\
One simple model that we will look at is the **True Flat Tax** model.  This is a constant tax rate on all gross income without any deductions.  I can only imagine that this model is what Ben Carson has in mind with his 10% flat tax proposal which he claims is based upon The Bible and his economists.  Ben Carson goes on to say that "eliminating the loopholes and deductions," the flat tax rate should fall in the 10-15% range, while Chris tells him that research by the Tax Policy Center suggests that a flat tax of 20-25% would be more reasonable to meet the current levels of tax revenue [^2].  Let's look at some numbers.

According to 2012 income tax [data](http://www.irs.gov/uac/SOI-Tax-Stats---Individual-Statistical-Tables-by-Size-of-Adjusted-Gross-Income#_grp1)[^3] published by the IRS, total U.S. tax revenue from individual returns was reported to be $1,188,027,000,000.  An approximation of the total gross income of Americans without any deductions or adjustments was reported as $9,100,131,000,000[^4].  A simple calculation shows that the flat tax needed to meet the 2012 individual tax revenue levels would be about 13%.  Not bad, it seems that Dr. Carson payed attention in algebra class as well as church.

Does a 13% true flat tax hurt anyone?  Well, let's look at an example of 17 year old Joe Blow who wants to work a summer job before his senior year in high school begins.  He may be getting paid $10 per hour and working 40 hours a week for 10 weeks.  That comes out to be a gross income of $4,000.  Under our current tax structure, he would claim the standard deduction of $6,300 leaving him with no taxable income, and thus he wouldn't pay any taxes.  However, under the true flat tax of 13% where deductions don't exist, he would end up paying $520 in taxes.  That's some serious gas money being taken out of Joe's pocket!

In Econ 101, we learned to assume that we as humans are rational decision makers.  Now, suppose Joe is faced with two options regarding his tax situation; paying no taxes due to a generous exemption and standard deduction or paying $520 in taxes.  Of course, being the rational human being that Joe is, he would choose to pay zero taxes.

```{r flatVSprog}
# flat tax and income info
flatRate <- .13
income <- seq(0,150000,by=100)
flatTax <- flatRate*income



# standard deductions
single <- 5950
married <- 11900
hOfHouse <- 8700

# ssTax for 2012 was 4.2% with max earnings of 110100
# medCare tax had no max at 1.45%
ssRate <- .042
medRate <- .0145


incomeRate <- function(income, deduction, rate, exemptions, ssRate, medRate){
    (income - (exemptions*3800) - deduction)*rate + income*(ssRate+medRate)
}

brackets <- c(0, 8700, 35350, 71350, 108725, 194175)
brkRates <- c(.1, .15, .25, .28, .33, .35)

# get bases for the progressive tax structure
progTaxes <- sapply(1:length(brackets), function(r) brkRates[r]*(brackets[r+1]-brackets[r]))
progTaxes <- progTaxes[-length(progTaxes)]


# in parallel, get values of tax for each do;;ar of gross income
#run 'foreach' in parallel
c1 <- makeCluster(4)
registerDoParallel(c1)
b1 <- foreach (i = iter(income), .combine=c, .multicombine=TRUE) %:%
    when(i >= 0 & i <= 8700) %dopar% incomeRate(i, single, rate=.1, exemptions=1,
                                                ssRate=.042, medRate=.0145)
b2 <- foreach (i = iter(income), .combine=c, .multicombine=TRUE) %:%
    when(i >= 8701 & i <= 35350) %dopar% (progTaxes[1] + 
                                              incomeRate(i-8701, single, rate=.15,
                                                         exemptions=1, ssRate=.042, medRate=.0145))
b3 <- foreach (i = iter(income), .combine=c, .multicombine=TRUE) %:%
    when(i >= 35351 & i <= 71350) %dopar% (sum(progTaxes[1:2]) +
                                               incomeRate(i-35351, single, rate=.25,
                                                          exemptions=1, ssRate=.042, medRate=.0145))
b4 <- foreach (i = iter(income), .combine=c, .multicombine=TRUE) %:%
    when(i >= 71351 & i <= 108725) %dopar% (sum(progTaxes[1:3]) + 
                                                incomeRate(i-71351, single, rate=.28,
                                                           exemptions=1, ssRate=.042,medRate=.0145))
b5 <- foreach (i = iter(income), .combine=c, .multicombine=TRUE) %:%
    when(i >= 108726 & i <= 110100) %dopar% (sum(progTaxes[1:4]) + 
                                                 incomeRate(i-108726, single, rate=.33,
                                                            exemptions=1, ssRate=.042, medRate=.0145))
b5.1 <- foreach (i = iter(income), .combine=c, .multicombine=TRUE) %:%
    when(i >= 110101 & i <= 194175) %dopar% (sum(progTaxes[1:4]) + 
                                                 incomeRate(i-108726, single, rate=.33,
                                                            exemptions=1, ssRate=0, medRate=.0145))
b6 <- foreach (i = iter(income), .combine=c, .multicombine=TRUE) %:%
    when(i >= 194176) %dopar% (sum(progTaxes[1:5]) + incomeRate(i-194176, single, rate=.33,
                                                                exemptions=1, ssRate=0, medRate=.0145))
iRate <- c(b1,b2,b3,b4,b5,b5.1,b6)

# pTax = progressive tax bracket structure
pTax <- ifelse(iRate <= 0, 0, iRate)
dfTFT <- data.frame(income, flatTax, pTax)
```

Comparing the 13% true flat tax to an oversimplified version of our progressive tax bracket structure in the figure below, we can see that individuals earning an income up to around $45,000 a year might not be too happy paying taxes under this True Flat Tax model.

```{r plotFlatProgLine}
linePlot <- ggplot(dfTFT, aes(x=income)) +
    geom_line(aes(y=flatTax, color="13% Flat Tax")) +
    geom_line(aes(y=pTax, color="Tax Bracket Model")) +
    scale_color_discrete(name = "Tax Models",
                         labels = c("13% Flat Tax", "Tax Bracket Model")) +
    ylab("Tax (dollars)") +
    xlab("Income (dollars)") +
    ggtitle("13% Flat Tax vs Progressive Tax Bracket Model \n
            Model Assumes Year 2012 Single Person Claiming One
            Exemption and Social Security/Medicare Tax") +
    theme(axis.text.x=element_text(size=15),
          axis.text.y=element_text(size=15),
          axis.title.x=element_text(size=18),
          axis.title.y=element_text(size=18),
          plot.title=element_text(size=20))
linePlot
```

```{r trueFlatStats}
##### Get the Flat tax Revenue for 13%
df$flatTaxRev <- .13*df$AGIlessDeficit

# upper range of flat tax revenue for incomes < 50000
upperFlatTaxRev <- sum(df$flatTaxRev[2:9])

# lower range of flat tax revenue for incomes < 50000
lowerFlatTaxRev <- sum(df$flatTaxRev[2:8])

# average flat tax per person
avgFlatTax <- upperFlatTaxRev/(sum(df$grossNumberOfReturns[2:9]))
df$avgFlatTaxPerPerson <- df$flatTaxRev/df$grossNumberOfReturns
```

In fact, using the comparison of these two tax models for our 2012 data, we estimate that the government would have received between $188 billion and $251 billion more in tax revenue under the 13% flat tax model than they would under our current progressive tax model for persons making less than $50,000 a year.  The following figure shows a side-by-side comparison of the 13% True Flat Tax and the current Progressive Tax revenues for each AGI level in our 2012 data.  Here we can see that the 13% True Flat Tax dominates in tax revenue up to around the $100-200K window where it starts to become dwarfed by the progressive tax revenue.  Hooray for the wealthy, and ouch if you're not quite living the American Dream yet.

```{r barPlot1}
##### create 'dodged' barplot of tax revenues for each income range for
## bot the flat tax and the 2012 progressive tax revenues
dfRevs <- melt(data.frame("Progressive Tax"=df$tax[2:length(df$tax)],
                          "13 Percent Flat Tax"=df$flatTaxRev[2:length(df$flatTaxRev)],
                          Income=df$AGI[2:length(df$AGI)]),
               variable.name="Model")


taxRevPlot <- ggplot(dfRevs, aes(Income, value, fill=Model)) +
    geom_bar(position="dodge", stat="identity") +
    ylab("Tax Revenue") +
    ggtitle("Actual 2012 Progressive Tax vs. 13% True Flat Tax") +
    scale_x_discrete(name="AGI Level", labels=df$AGI[2:length(df$AGI)],
                     limits=df$AGI[2:length(df$AGI)]) +
    scale_y_continuous(labels=dollar) +
    theme(axis.text.x=element_text(angle=90, size=13),
          axis.title.x=element_text(size=15),
          axis.title.y=element_text(size=15),
          plot.title=element_text(size=18))

taxRevPlot
```


```{r, echo = FALSE, cache=FALSE}
shinyApp(
  
  ui = fluidPage(
    selectInput("region", "Region:", 
                choices = colnames(WorldPhones)),
    plotOutput("phonePlot")
  ),
  
  server = function(input, output) {
    output$phonePlot <- renderPlot({
      barplot(WorldPhones[,input$region]*1000, 
              ylab = "Number of Telephones", xlab = "Year")
    })
  },
  
  options = list(height = 500)
)
```



## The Post Card Tax Return
>We should let taxes become so simple that they could be filled out on a postcard. <cite>Ted Cruz (R-TX), USA Today, October 19, 2014</cite>

>This simple 20% flat tax will allow Americans to file their taxes on a postcard. <cite>Rick Perry (R-TX), The Wall Street Journal, October 25, 2011</cite>

Does simplifying the federal tax code so much that we can file our tax return information on a few lines of a postcard sound good?  Well, it’s hardly an original idea by our presidential hopefuls.  Based on their statements, I would say that their ideas on tax reform are heavily borrowed from the idea that “tax forms really can fit on postcards” that Robert Hall and Alvin Rabushka of the Hoover Institution have proposed in their book [The Flat Tax](http://www.google.com/url?q=http%3A%2F%2Fwww.hooverpress.org%2Fproductdetails.cfm%3FPC%3D1274&sa=D&sntz=1&usg=AFQjCNESVCToAoD1idKhPnqPJjyogZoO1g) in 1985.

The Hall-Rabushka (HR) Flat Tax model is actually quite simple while still allowing for a generous exemption and standard deduction for taxpayers.  While HR models both an individual and a business tax (both of which are supposed to fit on a postcard) we will only look at the individual tax model here and save the business tax for another discussion another day.  The HR model can be summed up into just a few points:

* **Compensation** = Wages & Salary + Pensionions and Retirement Benefits
* **Deductions** = Standard Deduction - (Number of Exemptions \* Exemption dollar amount)
* **Taxable Income** = Compensation - Deductions
* **Tax** = flat tax percentage of taxable income    

Note that compensation does not include any income such as interest, dividends, capital gains, etc.  Just a simple earned income plus any private pensions you receive.  This model also wipes out all of those loopholes and deductions such as medical expenses, interest paid; just a simple standard deduction that should be quite generous and rise with the cost of living.  In fact, HR pointed out that the deductions for a family of four in 1995 would be $25,500.  Broken down these deductions are:

* in 1995 dollars
* $16,500 married filing jointly
* $9,500 single
* $14,000 single head of household
* $4,500 for each other dependent you claim (not including you or your spouse)

Now, I'm not sure how HR came up with these numbers, and they are slightly higher than the standard deduction amounts that the IRS published in 1995.  We will attempt to use these amounts adjusted with the Consumer Price Index to 2012 dollars to compare with our data.  The adjusted standard deduction amounts are:

* CPI adjusted for 2012 dollars and rounded to the nearest $100
* $24,900 married filing jointly
* $14,300 single
* $21,100 single head of household
* $6,800 for each other dependent you claim (not including you or your spouse)

Under these deductions, a family of four in 2012 would be allowed deductions of $38,500, which may or may not be a reasonable estimate?  For comparison, Rand Paul's (R-KY) is proprosing a similar tax model in which a family of four would be allowed $50,000 in deductions, but we'll get back to Rand Paul's proposal in just a bit.

Under this tax model, HR suggested that in 1995 the flat tax rate would be 19%. CAN WE ESTIMATE A FLAT TAX RATE BASED ON OUR 2012 DATA?

How would you fare under this tax system?  Go ahead and fill in your info in the boxes below and see how much taxes you would have to pay under this system.

[^1]: Wikipedia article on Flat Tax - https://en.wikipedia.org/wiki/Flat_tax
[^2]: An Interview with Fox News's Chris Wallace - http://mediamatters.org/video/2015/05/10/foxs-chris-wallace-exposes-truth-behind-ben-car/203587
[^3]: All Returns: Selected Income and Tax Items 2012 - http://www.irs.gov/uac/SOI-Tax-Stats---Individual-Statistical-Tables-by-Size-of-Adjusted-Gross-Income#_grp1
[^4]: Adjusted Gross Income less deficit - an approximation of gross income without deductions and adjustments. http://www.irs.gov/pub/irs-soi/12insec4.pdf
